---
title: "Installation for Introductory Bayesian Course"
output: html_document
---

If you haven't already done so, download the base distribution binary of R 3.1.2 for your platform from <http://cran.r-project.org/> and install using the default options. Next download and install the
most recent version of RStudio from <http://www.rstudio.com/products/rstudio/download/> using the default options. Then, download JAGS 3.4.0 from <http://sourceforge.net/projects/mcmc-jags/files/JAGS/> (right-click
and open this link in a new window) and install with the default options. 

To install all the required packages execute the following code at the R console.

```{r, eval = FALSE}
install.packages("devtools", quiet = TRUE)
install.packages("dplyr", quiet = TRUE)
install.packages("ggplot2", quiet = TRUE)
install.packages("scales", quiet = TRUE)

library(devtools)
install_github("poissonconsulting/tulip@v0.0.11")
install_github("poissonconsulting/datalist@v0.4")
install_github("poissonconsulting/juggler@v0.1.3")
install_github("poissonconsulting/jaggernaut@v2.2.3")
```

If everything was successful, when you run the following code
```{r, eval=FALSE, tidy=FALSE}
library(jaggernaut)
model <- jags_model("model {
 bLambda ~ dlnorm(0,10^-2)
 for (i in 1:length(x)) {
   x[i]~dpois(bLambda)
 }
}")

summary(jags_analysis (model, data = data.frame(x = rpois(100,1))))
```
you should see something like
```
Analysis converged (rhat:1)

Model1:

Dimensions:
samples  chains 
   1500       3 

Convergence:
rhat 
   1 

Estimates:
         estimate     lower    upper       sd error significance
bLambda 0.8771658 0.7151765 1.062476 0.088936    20            0
```